Dissertation draft

### 2023-07-22 Texfiles/Draft1.tex
Diss edits and code updates

\documentclass[12pt]{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[authoryear]{natbib}
\usepackage{enumerate}
\usepackage{booktabs}
\setlength{\parindent}{0pt}
\usepackage[parfill]{parskip}
\date{Lent 2023}
\title{Dissertation draft}
\author{Tobias Leigh-Wood}
\usepackage[letterpaper, margin=1.1in]{geometry}
\linespread{1.1}

@@

The main data set I use is the English Longitudinal Study of Ageing (ELSA) \cite{main_elsa_citation}. ELSA repeatedly interviews
individuals over the age of 50 and asks them a range of questions relating to their income and wealth as well as expectations about
the future. Importantly it also includes detailed information pensions including the type of pension that an individual holds when
working, thus I am able to distinguish between individuals who have defined benefit and defined contribution pensions. For ease of access
I use Harmonized ELSA \footnote{"This analysis uses data or information from the Harmonized ELSA dataset and Codebook, Version G.2 as of
  July 2021 developed by the Gateway to Global Aging Data. The development of the Harmonized ELSA was funded by the National
  Institute on Aging (R01 AG030153, RC2 AG036619, R03 AG043052). For more information,
  please refer to https://g2aging.org/.”} which ensures that variables are comparable across waves. Since this only includes a
subset of the questions in ELSA I also supplement it with variables taken directly from the data.

ELSA also includes questions on expenditures. In particular individuals are asked how much they consume on a range of broad categories.

% Include a summary stats table here of the individuals selected for the regression

I also use life tables from the UK's Office for National Statistics. These give us risk of death for each age group. I adjust them
to make death certain at age 110 as is common in the literature. I transform these so that I have risk of death conditional on being
a given age since this is what is used in the life cycle simulations.
%  include a table showing death probs

@@

\textbf{Add some other background to pensions in the UK. What does the Blundell etc have
  to say}

Successive governments in the UK have attempted to reform both the public and private pension system. Prior
to 1987 participation in private schemes was limited to employees of firms that had offered one, and there
were few alternatives to the public state pension or a define benefit scheme that a public sector employee
would offer. \textbf{Were there pension reforms in 1990s and 00s}

@@

@@

I regress demographic and financial characteristics of individuals on year of birth and the treatment dummy.
We can then see if the treatment groups differ on key characteristics such as financial wealth or retirement age.
For a regression discontinuity to be valid we need the treatment and control groups to be similar along
all other characteristics apart from the treatment variable.

In particular I run:
\begin{equation*}
  Y_{i} = \alpha + \beta Post_{i} + \gamma YOB + \kappa (Post_{i} \times YOB_{i}) + \epsilon_{i}
\end{equation*}

\input{tables/cov_balance.tex}
Table \ref{tab:cov_balance} shows these results. We can see ...

Another threat to validity is manipulation of the running variable. This is probably quite a threat in our case
as individuals could delay retirement and/or the purchase of an annuity.

@@

I use a regression discontinuity design where I compare the consumption of recent retirees after the policy reform
to recent retirees before the policy reform. The individuals who retired in 2014 and later were not forced to
annuitise their defined contribution pension. The key assumption implicit in regression discontinuities is
that nothing else changes at the time of the jump apart from the policy of interest. And that the policy occurs
without individuals predicting it. The policy change was widely seen as a surprise by the media and financial planners,
some of whom in fact complained that they had not been consulted enough. \textbf{How can I justify no other jumps at the jump}

Retirement year is the running variable and individuals are treated if retirement year is greater than 2013 and less than 2016.
I consumption of individuals up to 2 years into retirement so that the sample size is larger. So if someone retired in 2014 and
had consumption data in 2014 and 2016 I include both values. An individual is considered not treated if they retire before 2013 and after 2011.
One benefit of using ELSA is that it includes data on pension type for individuals who are working. Therefore, I can differentiate
between individuals who have a defined contribution pension and those who have a defined benefit contribution. To account for the fact
that the reform only impacted individuals who had accumulated defined contribution pension pots I interact the treatment dummy with
an indicator variable signalling whether the individual had ever held a DC pension pot.

The estimating equation is therefore

And $X_{it}$ is a set of controls including financial wealth etc

I also run the regression with a variety of consumption variables.

@@

          \item Add bit about DC/DB pensions in the intro. Also talk about heterogeneity across countries.
                Some countries want to move towards more annuitisation. This annual review is a good source of info
                \cite{banks_crawford_ar_2022}
        \end{enumerate}
  \item Lit review
  \item Models
  \item Empirical
        \begin{enumerate}
          \item Diff in diff
          \item RDD
          \item Matching?
          \item can I use anything I learnt in panel?
                In some sense the decision to annuitise is a discrete choice problem so I could use something from there.
        \end{enumerate}
  \item Conclusion
\end{itemize}

>>>>

\documentclass[12pt]{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[authoryear]{natbib}
\usepackage{enumerate}
\usepackage{booktabs}
\usepackage{pdflscape}
\setlength{\parindent}{0pt}
\usepackage[parfill]{parskip}
\date{Lent 2023}
\title{Dissertation draft}
\author{Tobias Leigh-Wood}
\usepackage[letterpaper, margin=1.1in]{geometry}
\linespread{1.1}

@@

The main data set I use is the English Longitudinal Study of Ageing (ELSA) \cite{main_elsa_citation}. ELSA repeatedly interviews
individuals over the age of 50 and asks them a range of questions relating to their income and wealth as well as expectations about
the future. One benefit of using ELSA is that it includes data on pension type for individuals who are working. Therefore,
I can differentiate between individuals who have a defined contribution pension and those who have a defined benefit contribution.
Importantly it also includes information on pension size that the Institute for Fiscal Studies calculated, this is only available
up until wave 5 at which point I use a real return of 3\% to predict forward the pension wealth variable until retirement.
ELSA also includes a measure of all non-housing financial wealth which I use in some specifications because of these issues with
pension wealth. Because of slight differences between years in some of the ELSA questions I use 'Harmonized ELSA'
\footnote{"This analysis uses data or information from the Harmonized ELSA dataset and Codebook, Version G.2 as of
  July 2021 developed by the Gateway to Global Aging Data. The development of the Harmonized ELSA was funded by the National
  Institute on Aging (R01 AG030153, RC2 AG036619, R03 AG043052). For more information,
  please refer to https://g2aging.org/.”} which ensures that variables are comparable across waves. Since this only includes a
subset of the questions in ELSA I also supplement it with variables taken directly from the data such as questions that deal
with life expectancy.

ELSA also includes questions on expenditures. In particular individuals are asked how much they consume on a range of broad
categories including in-house food consumption, out-of-house food consumption, leisure consumption, clothes consumption and
consumption on utility bills and rent. I use also use two questions on life expectancy that I discuss in more detail below.

I also use life tables from the UK's Office for National Statistics. These give us risk of death for each age group. I adjust them
to make death certain at age 110 as is common in the literature. I transform these so that I have risk of death conditional on being
a given age since this is what is used in the life cycle simulations.

@@

Successive governments in the UK have attempted to reform both the public and private pension system. Prior
to 1987 participation in private schemes was limited to employees of firms that had offered one, and there
were few alternatives to the public state pension or defined benefit scheme that a public sector employee
would offer. From 1989, individuals in the UK were allowed to open tax advantaged self-invested personal pensions
alongside any pension their employer offered. The largest reform prior to 2014 came in the 2004 Finance Act which
rationalised taxation rates on different types of pensions. However, defined contribution pension pots could still
only be used to buy an annuity after taking out a maximum of 25\% as a tax free lump sum withdrawal.

@@

Also of specific interest to the annuity market was the European Union's
'Equal Treatment in Goods and Services Directive' of 2004. This prohibited discrimination
in the provision and cost of good and services based on sex. Up until 2011 there was a clause that
stated insurers were allowed to charge different premiums if it was based on evidence that sex is
correlated with different amounts of risk. However, in March 2011 the European Court of Justice
ruled that insurers were not allowed to charge different amounts and gave them until December 2012
to implement the ruling. This change meant that annuity products could no longer be priced differently
for men and women in the UK. However, given this change was implemented two years before the pension
freedoms act, I can still identify the impact of the pensions freedom act on early retirement consumption.

@@

Table \ref{tab:sum_stats} shows various summary statistics for each variable of interest from ELSA and the ONS.
First note that the second retirement group is small with just 728 non missing observations for gender as opposed
to 941 individuals in the control group. The control group are less female, retired at a slightly younger age and expected to
retire slightly earlier. This could be because this period was affected by the increase in the state pension age for
women from age 60 in 2010 to age 65 in 2018. Since this change happened gradually and occured over the whole period
I do not expect it to influence the results. The DifferenceAge row tracks the difference between expected and actual
pension age.

As required, retirement year is between 2011 and 2013 for the treatment group and 2015 and 2017 for the control group.
However, as discussed above their interview date is after or the same as the retirement year because we are tracking
consumption in retirement.

The treatment group has higher financial wealth with a median of £65,000 at time of interview as opposed to £55,000
in the control group. Likewise, the second group are more likely to have held at defined contribution pension at some point
and have much more money in them. Both groups are equally likely to have a defined benefit pension with roughly 44\% of
individuals across both samples having a defined benefit pension. Individuals are much more likely to have a DB pension
than a DC pension which reflects the fact that retirees now are still more likely to have a DB pension although the number
with DC pensions are growing. Home ownership is roughly equal across groups although the second group have slightly more
housing wealth.

Both groups are objectively similarly long-lived according when using ONS life tables to calculate life expectancy
based on gender, age and the year the interview was carried out in. Subjective life expectancies are also similar
across groups, with individuals expecting to live another 21 years as opposed to the 24 that the ONS would expect
them to live. \textbf{Maybe add in prob of survival to age 85 or 90 or something}
\textbf{Also add in an indicator for living alone or living with someone else}

Unfortunately, some consumption data is missing for some individuals. The food categories have the least missing data
and the leisure consumption category has the most missing data. \textbf{Check why total consumption adds up to so much
  more than all the categories added together? Do they include rent and housing in consumption?}

Overall, the groups do appear to be slightly different with the treatment group being richer and slightly more likely
to have DC pension wealth. On key demographic characteristics though the groups are similar, the difference in retirement
age between the two groups can be mostly explained by the difference in expected retirement age meaning that individuals
have not on-mass decided to delay retirement and avoid annuitisation.

\begin{landscape}

\end{landscape}

As a further test for balance I regress demographic and financial characteristics of individuals on year of birth and the treatment dummy.
We can then see if the treatment groups differ on key characteristics such as financial wealth or retirement age.
For a regression discontinuity to be valid we need the treatment and control groups to be similar along
all other characteristics apart from the treatment variable.

In particular I run:
\begin{equation*}
  Y_{i} = \alpha + \beta Post_{i} + \gamma YOB_{i} + \kappa (Post_{i} \times YOB_{i}) + \epsilon_{i}
\end{equation*}

\input{tables/cov_balance.tex}
Table \ref{tab:cov_balance} shows these results. We can see that being in the treatment group, con

A key threat to validity is manipulation of the running variable. This is probably quite a threat in our case
as individuals could delay retirement and/or the purchase of an annuity until after the reform. To check whether
this is occuring we see whether the difference between real retirement age and the first expected retirement age
recorded in the data are different for those retiring before 2014 compared to those retiring after 2014. As shown
above in Table \ref{tab:sum_stats} this not the case. However, since we do not have data on the exact purchase
date of the annuity we cannot track whether annuity purchases were delayed because of this. However, we do still
relatively stable annuity demand before the policy reform. If individuals delayed annuity purchases in expectation
of the reform we would see steady declines in purchases before the reform was announced.

I use a regression discontinuity design where I compare the consumption of recent retirees after the policy reform
to recent retirees before the policy reform. The key assumption implicit in regression discontinuities is
that nothing else changes at the time of the jump apart from the policy of interest. And that the policy occurs
without individuals predicting it and changing their behaviour. As I have argued above the demographic information in the
data suggest that individuals did not delay retirement and that there was no delay in annuitisation as shown by the quick
and sudden decline in annuity purchases. Moreover, anecdotal evidence from media and business sources at the time of the
announcement were all surprised by the policy changes, with some financial planners even complaining that they had not been
briefed.

Retirement year is the running variable and individuals are treated if retirement year is greater than 2014 and less than
or equal to 2017. I use consumption data of individuals up to 2 years into retirement so that the sample size is larger.
So if someone retired in 2015 and had consumption data in 2015 and 2017 I include both values. An individual is considered
not treated if they retire before 2013 and after 2011.  To account for the fact
that the reform only impacted individuals who had accumulated defined contribution pension pots I interact the treatment dummy with
an indicator variable signalling whether the individual had ever held a DC pension pot.

Because of the differences in financial characteristics between the groups I add controls for financial wealth, housing wealth
and whether an individual owns their own home. I also run regression discontintuity with s

And $X_{it}$ is a set of controls.

@@

@@

          \item Add bit about DC/DB pensions in the intro. Also talk about heterogeneity across countries.
                Some countries want to move towards more annuitisation. This annual review is a good source of info
                \cite{banks_crawford_ar_2022}
          \item https://www.moneysavingexpert.com/savings/pension-freedom/
        \end{enumerate}
  \item Lit review
  \item Models
  \item Empirical
        \begin{enumerate}
          \item Diff in diff
          \item RDD
          \item Matching?
          \item can I use anything I learnt in panel?
                In some sense the decision to annuitise is a discrete choice problem so I could use something from there.
        \end{enumerate}
  \item Conclusion
\end{itemize}
